Joint maximum-likelihood of phylogenies and ancestral states is not consistent

### 2018-05-02 versions/response.v1.txt

> Associate Editor
> Editors’ comments to the author:
> All four reviewers agreed that the work is worth publishing and
> provides an important insight into the behavior of the joint
> ML/ancestral states estimation method. However, all reviewers asked
> for what amounts to major revisions to the text and presentation. You
> should especially focus on clarity and accessibility, provide the
> requested examples, and address suggestions on how the results could
> be even further improved.

Outline of revisions:
1.
2.

Some of the reviewers have expressed their opinion that we reduce our emphasis on mathematical development in exchange for a more numerical-optimization-based approach.
Our primary goal was to mathematically prove inconsistency of joint inference, which is what we have done.
As pointed out by reviewer 4, this answers an open question, which has special relevance given the recent work of Sagulenko et al.
We provide some numerical results to confirm our derivations and to provide a comparison.

As we now emphasize (TODO XXX) this work is not the final word on the joint inference question from an applied perspective.
Even if we were to greatly expand the sections doing numerical optimization, these results would only be for the binary symmetric model, which is never applied in practice.
An investigation of these aspects deserves another paper.

@@

> Finally, please pay VERY careful attention to improving your
> literature review (reviewer 4) and making a very clear case about how
> your work IMPROVES and EXTENDS previous results. And, please, do get
> rid of the Farris zone. The name "Farris" should not be attached to
> results that have to do with ML estimation, given the long and
> torturous efforts of cladists to undermine the entire line of inquiry.

We are grateful to reviewer 4 for pointing us to literature of which we were not aware.
We have greatly expanded ...

We have complied with this request regarding terminology, despite it going against current practice.
Indeed, Felsenstein's book has a subsection entitled "The Farris zone" spanning two pages (pp 120-121), a figure entitled "A tree in the Farris zone.", and a table with a caption "Properties of the Felsenstein and Farris zones...".
The term "Farris zone" has almost 7 times as many citations as the term "inverse Felsenstein zone" on Google Scholar.

@@

> Reviewer: 1

> Comments to the Author
> The authors show that when ancestral states are considered as
> parameters, there are zones where topologies and edge-lengths are
> inconsistently estimated. As is evidenced by any ML package, it is
> widely accepted that ancestral states should not be parameters; the
> large number of parameters suggests the approach will be problematic.
> Still, to the best of my knowledge, it has not been shown that
> treating ancestral states as parameters actually leads to poor
> performance, or how poor. Moreover, the authors point to Sagulenko et
> al (2018) - now published in Virus Evolution - as an example
> implementation. I found the paper interesting and I believe a more
> succinct presentation will be of interest.

> A main concern with the paper is that the shaded regions of
> topological inconsistency or poor estimation in Figs 2-4, are only
> subsets of the actual regions where topological inconsistency or poor
> estimation occur; and yet determining the actual regions seems
> feasible. The contrast between Fig 5 and 4 illustrates how far off the
> subsets are in giving the actual regions where estimation is
> problematic. The white region in Fig 5 is the actual region where hat
> w=1 whereas the shaded region of Fig 3 is the subset of the actual
> region that the authors were able to determine through analytical
> arguments. For most readers, out of Fig 3 and Fig 5, only Fig 5 is of
> interest.

As described above, our intent is to mathematically establish inconsistency and poor estimation.
This must be done with analytic arguments rather than optimization.
We could have only shown this bad behavior in a very small area to establish that it happens.
In contrast to this reviewer, we believe that Fig 5 is optional to this line of argument, and we provide it just for interest.
A true investigation into the practical implications of our work will require another paper.

@@

> The other main concern that I have is with the restriction that
> t=(x,y,x,y,w) in estimation rather than t=(t_1,...,t_5). On the one
> hand, usually one expects that incorporating knowledge about the true
> underlying tree (t_1=t_3 & t_2=t_4 in the present case) should be
> helpful for a good estimation procedure. Thus the fact that treating
> ancestral states as parameters gives inconsistent topological
> estimation in this setting demonstrates that the approach is
> problematic. On the other hand, allowing t=(t_1,...,t_5) is what would
> be done in any practical implementation and it seems quite possible to
> me that results would be radically different in this case. For
> instance, from Fig 2, the generating tree with all equal edge-lengths
> (y*=x*=0.2) provides a case where topology tau_2 is inconsistently
> estimated with the restriction that estimated t=(x,y,x,y,w). Because
> of the symmetry of the problem, I am doubtful y*=x*=0.2 would be a
> point of inconsistency if t=(t_1,...,t_5) were unrestricted in
> estimation. The issue deserves discussion because I don't think it is
> transparent.

TODO: make the argument more clear in the main text.

> I don't think the restricted case of p8-9 will be of interest to
> readers. Only the general case.

%EM I think we could move this to the appendix to throw this reviewer a bone.

@@

> Reviewer: 2

> Comments to the Author
> In this manuscript  the authors show that joint inference of phylogeny
> and ancestral characters is not consistent in general by providing a
> counter-example, in which as the number of sites grows, they show that
> the wrong tree topology can have a higher likelihood than the correct
> topology under the joint inference model.
> The result is important for the community to understand, in particular
> because TreeTime (the joint inference method) is used in high-profile
> analyses on Ebola, influenza and other viruses on the nextflu /
> nextstrain websites. These are beautiful and accessible sites aiming
> to make analyses on viruses open to exploration by the community, but
> this comes at the cost of suppressing the details of the analysis. If
> indeed there are deep statistical issues or limitations it is
> important that the community knows what these are and what their
> implications are.

> Accordingly my comments are focussed on where I think the manuscript
> could be more clear (and where I did not quite follow the logic). My
> major suggestion would be to go even further towards demonstrating in
> a simple example how the joint inference can be wrong – for example,
> providing a toy multiple sequence alignment or even one of some
> influenza HA sequences – you could find 2 sets of 4 closely-related
> sequences that have the Farris and Felsenstein topologies, maybe? And
> demonstrate that the inference gives the wrong result by explicitly
> computing the likelihood.

We agree that understanding the implications of our derivation is a very interesting avenue for future work.
However, to do such a project justice would require a separate paper.
We are sure that once our paper demonstrates this inconsistency, others will quickly follow to write such a paper.

@@

> Also, laying out the theory a little bit more step-by-step so that a
> general quantitative reader doesn't struggle too much to follow would
> help (suggestions below).

Thank you very much for reading the derivations and for your suggestions. As noted inline we have revised the text accordingly.

@@

> I know that it's not feasible to explicitly compute (analytically)
> these likelihoods for bigger trees. But one of the main motivations
> and applications for tools like TreeTime is surely that the number of
> sequences grows (not the number of sites, which is fixed by the viral
> sequence length in the case of viruses). What happens when the number
> of tips grows? Are there more opportunities for the kind of issue that
> gives the wrong value for w? Many influenza trees have many 0 (or near
> 0) internal branch lengths, and it would be interesting to understand
> the implications for the ancestral state reconstructions (which in
> turn are used to infer spatial spread, transmission etc).

%EM David-- does it seem like we could extend these results to bigger trees containing a 4 taxon tree?

@@

> Reviewer: 3

> Comments to the Author
> Shaw and Matsen show that a joint maximum-likelihood inference of
> phylogenies and ancestral states is not consistent when branch lengths
> are long. We have recently proposed such a joint estimation for
> sequences that recently diverged and are hence similar (Sagulenko et
> al). The results presented here are anticipated in Sagulenko et al
> "For more divergent sequences, however, subleading states of internal
> nodes make a substantial contribution and the iterative optimization
> will underestimate the branch lengths."
> and
> "The iterative optimization procedures are not expected to be accurate
> for trees were many sites are saturated. In scenarios with extensive
> uncertainty of ancestral states and tree topology, convergence of the
> iterative steps cannot be guaranteed."

> In this light, it should be pointed out that joint estimation
> procedure was explicitly suggested as an approximation and hence never
> expected to be consistent in a strict sense. Furthermore, it was
> suggested as an approximation when branch lengths are short -- the
> opposite of the problematic regions identified here.
> But systematic analysis of the behavior of this approximation is
> useful and I believe the work is correct (but haven't checked the
> proofs). The work by Shaw and Matsen suggests the radius of
> convergence of this procedure is somewhere around branch length of
> order 10-20%, which is consistent with results reported by Sagulenko
> et al.

Thank you for your thoughts, and we have worked hard to not oversell our results.
TODO we should make some changes to at least say that we've further clarified the implications of our results.
The TreeTime paper does present joint optimization as an approximate approach, and described that it is for short branch lengths.
However, in our assessment the word "approximate" doesn't completely describe a situation in which there is significant bias.
The TreeTime paper says that its use is appropriate when "ancestral sequence inference has little ambiguity" but even in these cases the bias in branch lengths persists.

> At the same time, these results are not particularly surprising.
> Assigning ML ancestral states will always result in shorter branches.
> For short branches, this correction is linear in branch length, for
> longer branches higher order effects start to contribute. If the
> typical branch length around a bifurcating node is x<<1, a fraction ~x
> of the sites mutated along one of the three adjacent branches, while a
> fraction ~x^2 mutated on two branches and not the third. The latter
> case is ignored in this joint inference procedure. Not having to sum
> over these subleading states is what makes joint inference faster, but
> such truncation will first lead to inaccuracies and at longer
> distances to qualitatively different solutions (like Taylor expanding
> 1/(1-x) ).
>
> There are other obvious cases where joint inference will fail. For
> sequences with a heavily biased in nucleotide composition (say biased
> towards A). For sufficiently long terminal branches, all ancestral
> states will be estimated as all 'A' and subsequent branch length
> estimate will shrink all internal branches to zero. This happens for
> the very same reason that the most likely outcome of multiple flips of
> a biased coin is all heads -- a very atypical yet most likely outcome.
>
> Currently, the work largely shows that areas of inconsistencies exist.
> It would be useful if the authors could present Taylor expansions for
> the ML branch length estimates around x=y=w=1. This would more
> explicitly show the inaccuracies accumulate with increasing branch
> length and how the inaccuracies depend on asymmetries in the tree. To
> me, this would be much more useful than to know that joint inference
> will fail if branches are long.

This is an interesting suggestion but is different than our goal, which was to establish inconsistency.
TODO describe why Taylor isn't a straightforward thing.
TODO refer to idea of applied paper again?

@@

>  - line 50: overestimates branch lengths? It overestimates fidelity ->
>  underestimates branch length.  As expected.  Again line 249. Please
>  check. The back and forth between fidelity and branch length can be
>  confusing.

>  - Fig 5: it is surprising that \hat{w} seems completely flat as a
>  function of x* until it jumps discontinuously to 0. Why not include
>  the analytical bound in Fig 5?

>  - It would be nice if the abstract pointed out that these
>  inconsistencies arise when branch lengths are long and that the
>  approximate joint maximization procedure was suggested for precisely
>  the opposite limit. Similar at the beginning of the discussion, I
>  think it would be more appropriate if this would say smth along the
>  lines "Sagulenko et al suggested an approximation for short
>  branches... we show that this doesn't work for long branches..."

TODO we should do some change, but make it clear that it's consistently biased even for short branches.

@@

> Comments to the Author
> Review of Shaw, D. and Matsen, F.  "Joint maximum-likelihood of
> phylogenies and ancestral states is not consistent"

> Shaw and Matsen demonstrate that the approach of jointly estimating
> ancestral states and tree topologies (or branch lengths on a fixed
> topology) is not statistically consistent.  I believe that their
> results are correct, and that their result is important to
> disseminate.  The paper is clearly written, given the complexity of
> the material, and I have no reservation about recommending publication
> virtually as is.

> However, although my original intent was to carefully work through the
> math (including the derivations and proofs in the appendix), I have
> not found the time to do so, and because of the lateness of my review
> at this point, I will provide just a few comments in the hope that
> they will be useful.  Hopefully, other reviewers were able to check
> the correctness of the mathematical results more carefully.

Thank you for your assessment.
We have worked hard in our initial submission and in our revision to write out the argument in great (indeed, excruciating) detail, and we would appreciate it if you could take a look at the revision.

@@

> The only real criticism I have of the paper is that the authors could
> have done a better job in connecting their new work to previous
> results.  For example, Goldman (1990, Syst. Biol. 39:345-361) showed
> that joint estimation of topology and ancestral states by maximum
> likelihood was equivalent to maximum parsimony analysis for the same
> 2-state model considered by Shaw and Matsen when all branches are
> forced to have the same length.  They noted the statistical
> inconsistency of this approach, which was necessarily true because
> parsimony was already known to be inconsistent (the famous result of
> Felsenstein 1978).  Shaw and Matsen extend this result to the case of
> unequal branch lengths, which while not particularly surprising, is
> nonetheless an important extension.  The authors should also cite key
> relevant earlier work on the identifiability ot the tree topology with
> "classical" maximum likelihood phylogenetic estimation (Allman et al
> 2008, Adv Appl Prob 40:229–249; Chai and Housworth 2011, Syst Biol
> 60:713–718) in addition to crediting Roychoudhury 2015 for
> establishing ML consistency.

We are very grateful to this reviewer for pointing out the Goldman paper.
We were not aware of this paper and now have a paragraph describing it and comparing to our work.
It's also a good idea to cite additional papers about identifiability, and we have done so.

%EM David, heh, can you point me to the place where the equal branch length result was done?
%EM Also, I think you said that he says that the varying branch length thing is an open question-- it would be great to quote that.

@@

> Finally, let me just add that I hate the term "Farris Zone".  The
> "Felsenstein Zone" was a name applied to Felsenstein's important
> demonstration of a tree and branch-length configuration where
> parsimony would be an inconsistent estimator of the phylogeny.  On the
> contrary, Farris had nothing to do with the "Farris zone"; it was
> simply a name given by Mark Siddall to a different configuration in
> which parsimony outperforms likelihood, thinking incorrectly that
> likelihood was inconsistent in this zone.  Swofford et al. (2001,
> cited by the authors) referred to this zone as the "inverse
> Felsenstein zone", which I think is a better name for it.

As described at the beginning, we have made this change, thanks.

### 2018-05-02 appendix.tex
Minor changes and suggestions; draft response.

Since we have a finite character alphabet, for a given column $i$ there are a finite number of possible assignments of characters to tips $\alignmentColumn_i$ or internal nodes $\ancestralStateColumn_i$.
Take the tip labels of $\tau$ to be $\{1,\ldots,\nSiteRows\}$.
For likelihood calculation under the binary symmetric model, we describe a given $\alignmentColumn_i$ as a subset of indices $\siteSplit\subseteq\siteSplitSet:=\{1,\ldots,\nSiteRows-1\}$ with equivalent characters, commonly called a ``site split.''
We define the site split $\siteSplit$ for a $\alignmentColumn_i$ as simply $\alignmentColumn_i$ if the label $\nSiteRows$ is not in $\alignmentColumn_i$, and as its complement otherwise.
Taking such a complement simplifies but does not change the result of likelihood computation because the probability of observing a particular collection of binary characters is equivalent to the probability of its complement under the binary symmetric model.

@@

We enumerate the site splits $\siteSplit_j$ of which there are $\nSiteSplits=|\mathcal{P}(\siteSplitSet)|$ in total where $\mathcal{P}$ denotes the power set.
Similarly we enumerate ancestral splits $\ancestralSplit_k$ of which there are $\nAncestralSplits=|\mathcal{P}(\ancestralSplitSet)|$ in total.

@@

    &\qquad\qquad + p_{23}         \cdot\log((1+x)(1-x)(1+w)(1+y)(1-y))
+ p_{123}        \cdot\log((1+x)^2   (1+w)(1+y)(1-y)) \\

@@

The other two possible ancestral state splits for the likelihood of the site split $\{1,3\}$ admit similar simplifications.
The upper bound in \eqref{eq:farris-upper-bound} is for the ancestral state split $\{1\}$.
%EM Let's be more specific, giving an equation number, rather than just saying "the upper bound" here.
For the ancestral state split $\emptyset$ we see the upper bound will be the same except we lose a $(1+x)^2$ term, gain a $(1-x)^2$ term, lose a $(1-w)$ term, and gain a $(1+w)$ term.
In this case, all terms involving $w$ simplify to $\log(1+w)$, maximized at $\hat{w}=1$.
For the constants above, $a_{1}$ is the multiplier for the $\log(1+x)$ term and $a_{2}$ for the $\log(1-x)$ term meaning that exchanging the above quadratic $x$ terms yields new constants
\begin{equation}
    \begin{aligned}
        a_{1}' &= a_{1}-2p_{13}, \\
        a_{2}' &= a_{2}+2p_{13}, \\
    \end{aligned}
    \label{eq:a_const_prime_x}
\end{equation}
for site split $\emptyset$ and
\begin{equation}
    \begin{aligned}
        a_{3}' &= a_{3}-2p_{13}, \\
        a_{4}' &= a_{4}+2p_{13}, \\
    \end{aligned}
    \label{eq:a_const_prime_y}
\end{equation}
for site split $\{1,2\}$.
To bound the case of $\ancestralSplitPartition_j(\tau_1, t) = \emptyset$,
\begin{align*}
&    \ell_{\tau_1,t^*}(\tau_1, t) \\
&\qquad\le      p_{\emptyset}  \cdot\log(2)
+ a_{1}'\cdot\log\frac{2a_{1}'}{a_{1}'+a_{2}'}
+ a_{2}'\cdot\log\frac{2a_{2}'}{a_{1}'+a_{2}'} \\
&\qquad\qquad + a_{3}\cdot\log\frac{2a_{3}}{a_{3}+a_{4}}
+ a_{4}\cdot\log\frac{2a_{4}}{a_{3}+a_{4}}
+ \log(2) \\
&\qquad := C^{2}_{\tau_1,\tau_1}(x^*, y^*),
\end{align*}
and, for the case of $\ancestralSplitPartition_j(\tau_1, t) = \{1,2\}$,
\begin{align*}
&    \ell_{\tau_1,t^*}(\tau_1, t) \\
&\qquad\le      p_{\emptyset}  \cdot\log(2)
+ a_{1}\cdot\log\frac{2a_{1}}{a_{1}+a_{2}}
+ a_{2}\cdot\log\frac{2a_{2}}{a_{1}+a_{2}} \\
&\qquad\qquad+ a_{3}'\cdot\log\frac{2a_{3}'}{a_{3}'+a_{4}'}
+ a_{4}'\cdot\log\frac{2a_{4}'}{a_{3}'+a_{4}'}
+ \log(2) \\
&\qquad := C^{3}_{\tau_1,\tau_1}(x^*, y^*).
\end{align*}

We now need to check that
$$
\hat{x} = \frac{a_{1}'-a_{2}'}{a_{1}'+a_{2}'}, \ \hat{y} = \frac{a_{3}'-a_{4}'}{a_{3}'+a_{4}'}
$$
are both between zero and one.
Using a similar argument as that where $\ancestralSplitPartition_j(\tau_1, t) = \{1,3\}$, we only need to show $a_1' \ge a_2'$, which is true if $3-2p_{13} \ge 3\tilde{p}+2p_{13}$.
Some rearranging and the fact that $0 \le \tilde{p} \le 1/2$ shows this is equivalent to showing $p_{13} \le 3/8$.
Using Table~\ref{tab:sitepatprob} and $t^*$,
$$
p_{13} = \frac{1}{8}\left(1 + (x^*)^2 + (y^*)^2 - 4x^*(y^*)^2 + (x^*)^2(y^*)^2\right),
$$
meaning we are interested in whether
\begin{equation}
1 + (x^*)^2 + (y^*)^2 - 4x^*(y^*)^2 + (x^*)^2(y^*)^2 \le 3.
\label{eq:generating_ineq}
\end{equation}
We see that
\begin{equation}
1 + (x^*)^2 + (y^*)^2 - 4x^*(y^*)^2 + (x^*)^2(y^*)^2 \le 3 \iff (y^*)^2\left(1 - 4x^* + (x^*)^2\right) \le 2 - (x^*)^2
\label{eq:lenny}
\end{equation}
and that
\begin{align*}
(y^*)^2\left(1 - 4x^* + (x^*)^2\right) &\le (y^*)^2\left(1 - 2x^* + (x^*)^2\right) \\
                                       &=   (y^*)^2\left(1 - x^*\right)^2 \\
                                       &\le        \left(1 - x^*\right)^2 \\
                                       &\le              2 - (x^*)^2,
\end{align*}
with the last inequality holding if $0 \le x^* \le 1$, showing that \eqref{eq:lenny} holds.
Thus, $p_{13} \le 3/8$ and our $\hat{x}$ and $\hat{y}$ are valid.

>>>>

Since we have a finite character alphabet, for a given column $i$ there are a finite number of possible assignments of characters to tips $\alignmentColumn_i$ or internal nodes $\ancestralStateColumn_i$.
Take the tip labels of $\tau$ to be $\{1,\ldots,\nSiteRows\}$.
For likelihood calculation under the binary symmetric model, we describe a given $\alignmentColumn_i$ as a subset of indices $\siteSplit\subseteq\siteSplitSet:=\{1,\ldots,\nSiteRows-1\}$ with equivalent characters, commonly called a ``site split.''
We define the site split $\siteSplit$ for a $\alignmentColumn_i$ as simply $\alignmentColumn_i$ if the label $\nSiteRows$ is not in $\alignmentColumn_i$, and as its complement otherwise.
Taking such a complement simplifies but does not change the result of likelihood computation because the probability of observing a particular collection of binary characters is equivalent to the probability of its complement under the binary symmetric model.
%EM "ancestral state splits" are an important item in the development but don't get the same level of definition as "site splits". Let's define them here.

@@

We enumerate the site splits $\siteSplit_j$ of which there are $\nSiteSplits=|\mathcal{P}(\siteSplitSet)|$ in total where $\mathcal{P}$ denotes the power set.
%EM I think that we should either use "ancestral splits" everywhere, or "ancestral state splits."
Similarly we enumerate ancestral splits $\ancestralSplit_k$ of which there are $\nAncestralSplits=|\mathcal{P}(\ancestralSplitSet)|$ in total.

@@

    &\qquad\qquad + p_{23}         \cdot\log((1+x)(1-x)(1+w)(1+y)(1-y)) \\
    &\qquad\qquad + p_{123}        \cdot\log((1+x)^2   (1+w)(1+y)(1-y)) \\

@@

The other two possible ancestral state splits for the likelihood of the site split $\{1,3\}$ admit similar simplifications, as we now describe.
The upper bound in \eqref{eq:farris-upper-bound} is for the ancestral state split $\{1\}$.
%EM Let's be more specific, giving an equation number, rather than just saying "the upper bound" here.
For the ancestral state split $\emptyset$ we see the upper bound will be the same except we lose a $(1+x)^2$ term, gain a $(1-x)^2$ term, lose a $(1-w)$ term, and gain a $(1+w)$ term.
In this case, all terms involving $w$ simplify to $\log(1+w)$, which is maximized at $\hat{w}=1$.
For the constants above, $a_{1}$ is the multiplier for the $\log(1+x)$ term and $a_{2}$ for the $\log(1-x)$ term meaning that exchanging the above quadratic $x$ terms yields new constants
\begin{equation}
    \begin{aligned}
        a_{1}' &= a_{1}-2p_{13}, \\
        a_{2}' &= a_{2}+2p_{13}, \\
    \end{aligned}
    \label{eq:a_const_prime_x}
\end{equation}
for site split $\emptyset$ and
\begin{equation}
    \begin{aligned}
        a_{3}' &= a_{3}-2p_{13}, \\
        a_{4}' &= a_{4}+2p_{13}, \\
    \end{aligned}
    \label{eq:a_const_prime_y}
\end{equation}
for site split $\{1,2\}$.
To bound the case of ancestral state split $\ancestralSplitPartition_j(\tau_1, t) = \emptyset$,
\begin{align*}
&    \ell_{\tau_1,t^*}(\tau_1, t) \\
&\qquad\le      p_{\emptyset}  \cdot\log(2)
+ a_{1}'\cdot\log\frac{2a_{1}'}{a_{1}'+a_{2}'}
+ a_{2}'\cdot\log\frac{2a_{2}'}{a_{1}'+a_{2}'} \\
&\qquad\qquad + a_{3}\cdot\log\frac{2a_{3}}{a_{3}+a_{4}}
+ a_{4}\cdot\log\frac{2a_{4}}{a_{3}+a_{4}}
+ \log(2) \\
&\qquad := C^{2}_{\tau_1,\tau_1}(x^*, y^*),
\end{align*}
and, for the case of ancestral state split $\ancestralSplitPartition_j(\tau_1, t) = \{1,2\}$,
\begin{align*}
&    \ell_{\tau_1,t^*}(\tau_1, t) \\
&\qquad\le      p_{\emptyset}  \cdot\log(2)
+ a_{1}\cdot\log\frac{2a_{1}}{a_{1}+a_{2}}
+ a_{2}\cdot\log\frac{2a_{2}}{a_{1}+a_{2}} \\
&\qquad\qquad+ a_{3}'\cdot\log\frac{2a_{3}'}{a_{3}'+a_{4}'}
+ a_{4}'\cdot\log\frac{2a_{4}'}{a_{3}'+a_{4}'}
+ \log(2) \\
&\qquad := C^{3}_{\tau_1,\tau_1}(x^*, y^*).
\end{align*}

We now need to check that
$$
\hat{x} = \frac{a_{1}'-a_{2}'}{a_{1}'+a_{2}'}, \ \hat{y} = \frac{a_{3}'-a_{4}'}{a_{3}'+a_{4}'}
$$
are both between zero and one.
Using a similar argument as that where $\ancestralSplitPartition_j(\tau_1, t) = \{1,3\}$, we only need to show $a_1' \ge a_2'$, which is true if $3-2p_{13} \ge 3\tilde{p}+2p_{13}$.
Some rearranging and the fact that $0 \le \tilde{p} \le 1/2$ shows this is equivalent to showing $p_{13} \le 3/8$.
%EM Using what about t^*?
Using Table~\ref{tab:sitepatprob} and $t^*$,
$$
p_{13} = \frac{1}{8}\left(1 + (x^*)^2 + (y^*)^2 - 4x^*(y^*)^2 + (x^*)^2(y^*)^2\right),
$$
meaning we are interested in whether
\begin{equation}
1 + (x^*)^2 + (y^*)^2 - 4x^*(y^*)^2 + (x^*)^2(y^*)^2 \le 3.
\label{eq:generating_ineq}
\end{equation}
We see that
\begin{equation}
\begin{split}
& 1 + (x^*)^2 + (y^*)^2 - 4x^*(y^*)^2 + (x^*)^2(y^*)^2 \le 3 \\
& \qquad \qquad \qquad \qquad \iff (y^*)^2\left(1 - 4x^* + (x^*)^2\right) \le 2 - (x^*)^2
\end{split}
\label{eq:lenny}
\end{equation}
and that
\begin{align*}
(y^*)^2\left(1 - 4x^* + (x^*)^2\right) &\le (y^*)^2\left(1 - 2x^* + (x^*)^2\right) \\
                                       &=   (y^*)^2\left(1 - x^*\right)^2 \\
                                       &\le        \left(1 - x^*\right)^2 \\
                                       &\le              2 - (x^*)^2,
\end{align*}
with the last inequality holding if $0 \le x^* \le 1$, showing that \eqref{eq:lenny} holds.
Thus, $p_{13} \le 3/8$ and our $\hat{x}$ and $\hat{y}$ are valid.
